Topographic stresses and the Wenchuan earthquake

### 2014-05-29 manuscript/styron_hetland_wenchuan_topo_manuscript.tex
added hetland's revisions to manuscript tex

%\usepackage[unicode=true]{hyperref}
%\hypersetup{breaklinks=true,
%            bookmarks=true,
%            pdfauthor={},
%            pdftitle={Topographic stresses and the Wenchuan earthquake},
%            colorlinks=true,
%            citecolor=blue,
%            urlcolor=blue,
%            linkcolor=magenta,
%            pdfborder={0 0 0}}
%\urlstyle{same}

@@

Stress is of fundamental importance to many processes in the earth. Both
the isotropic (pressure) and anisotropic (shear) components of stress
exert control on the deformation state of the earth at any point in the
brittle regime. However, unlike other fundamental quantities such as
temperature, stress is typically difficult to measure \emph{in situ},
without drilling-based techniques. Therefore, stress is often treated in
a semi-quantitative manner, with an emphasis on directions and relative
magnitudes of the principal stresses, either locally or regionally
\citep[e.g.,][]{angelier1994}. These estimates of stress are commonly
derived from strain, for example from studies of earthquake focal
mechanisms \citep[e.g.,][]{michael1987} or of geological fault slip data
\citep[e.g.,][]{reches1987}.

@@

In this work, we investigate the effects of topographic stresses on the
faults that ruptured in the 2008 \emph{M}7.9 Wenchuan, China earthquake.
This earthquake is an ideal candidate for study because it occurred at
the base of the Longmen Shan, one of the largest and steepest
escarpments on Earth (Figure~\ref{fig:lms_map}), and has a well-studied
coseismic slip distribution characterized by significant along-strike
variations in coseismic slip. Additionally, because the earthquake
occurred after \textasciitilde{}2000 years of seismic quiescence,
postseismic stresses in the lithosphere are likely to be negligible,
suggesting that the stress state on the faults at the time of rupture
can be approximated as the sum of topographic, lithostatic, and
accumulated tectonic stresses. We then use the Mohr-Coulomb failure
criterion to bracket the tectonic stress field, pore fluid pressure, and
static friction of the fault.

@@

Aspects of topographic stresses and their relevance to tectonics have
been studied for some time. \citet{jeffreys1924} noted that the
presence of high mountains is evidence that the Earth's crust can
support significant heterogeneous differential stresses over long time
periods. \citet{dalmayrac1981} and \citet{molnar1988} discussed how 
extensional deformation in
the high parts of orogens temporally coincident with contractional
deformation at the low-elevation margins of the orogen may be explained
by a spatially invariant, depth-integrated horizontal compressive stress
with spatially varying vertical stresses caused by topography.
\citet{richardson1994} exploited this
relationship in the central Andes to estimate horizontal tectonic
stresses through finite element modeling, and more recently 
\citet{copley2009} 
performed similar work in Albania. \citet{bollinger2004} showed how 
increased normal stress on the Main
Himalayan thrust due to loading of the high Himalayan massifs locally
suppressed microseismicity and increased fault locking. Meade and Conrad
\citet{meade2008} demonstrated that the increased weight of the
uplifting Andes influenced the Nazca-South America convergence rate.

Additionally, topographic stresses play a role in common models of
orogenic dynamics. For example, topographic loading is central to thin
viscous sheet models of lithospheric deformation \citep[e.g.,][]{birdpiper80,
flesch2010gpe}. Critical taper models for thrust or extensional
wedges \citep[e.g.,][]{dahlen1990, xiao1991} incorporate both variation in
vertical stress due to changing elevation as well as a shear stress
contributed by the slope of the wedge surface. Critical taper models
also incorporate the idea that in a growing wedge, progressively
increasing topographic stresses may eventually prevent continued slip on
a given fault plane, and strain will instead be transferred to the toe
of the thrust wedge where the stress state is more favorable. Of
particular relevance to this work are `fixed boundary' models of
gravitational collapse and spreading \citep{rey2001}, where an excess of
gravitational potential energy associated with the high topography and
thickened crust of eastern Tibet causes a transfer of rock to the
foreland through horizontal contraction at the rangefront \citep[e.g.,][]
{dewey1988, liuyang2003, copleymckenzie2007}. This process may be
aided by a weak (sub)horizontal structure at depth (such as a shear zone
or weak crustal channel) that is capable of transferring the vertical
and radial stresses from topography throughout an orogen to its margins,
leading to contraction there \citep[e.g.,][]{clark2005, burchfiel2008,
fleschbendick2012}.

The contributions of variable topography to the full stress field in the
elastic upper crust has been studied on smaller spatial scales. 
\citet{mctiguemei1981} and
\citet{savageswolfs1986} investigated the stress components from long,
symmetric ridges and showed how horizontal tension is induced under
ridge crests and horizontal compression is induced under valleys, mostly
due to shear stresses generated by slopes. Work in this vein was
continued by \citet{millerdunne1996} and \citet{martel2006}, who have 
focused on shallow rock fracturing
resulting from topographic stresses. \citet{liuzoback1992} investigated 
whether the topographic stresses
generated by the mountains around Cajon Pass (California, USA)
contributed to the observed left-lateral shear stress on a shallow
portion of the right-lateral San Andreas fault. In their study, they
developed methods for calculating the three dimensional elastic stress
tensor field arbitrary topography, where previous solutions were limited
to two dimensions and required topography to be mathematically-defined
(e.g., sinusoidal).

@@

Topographic stresses are only one component of the total stress field in
the crust \citep{molnar1988}. Coseismic slip in the Wenchaun earthquake
is due to the total accumulated stress on the faults, which also
includes components from lithostatic and tectonic stresses. (In the
present study, we do not consider stresses due to flexure \citep[e.g.,][]
{luttrell2007} or buoyancy \citep[e.g.,][]{luttrell2011}.) By quantifying
both the topographic and lithostatic stresses, we can use coseismic slip
models to solve for the tectonic stress assuming that (1) the total
shear stress on the fault is in the direction of coseismic slip \citep[e.g.,]
[]{angelier1994} and (2) the fault is at Mohr-Coulomb failure everywhere
that is slipped. If topographic stresses are significant and produce
shear in the direction of fault slip, then for given values of static
friction and pore fluid pressure, we can calculate the amount of
tectonic stress that can be added to the ambient stress field before the
faults should rupture; given limited acceptable ranges for friction and
fluid pressure, we are essentially able to place maximum constraints on
tectonic stress. Alternately, if topographic stresses work against
coseismic slip, for given friction and fluid pressures we can estimate
the minimum magnitudes of tectonic stresses necessary to overcome shear
and frictional resistance to slip. In a scenario with complex faulting
and topography, it may be possible to put tight bounds on minimum and
maximum magnitudes and directions of tectonic stresses. To account for
the non-uniqueness of the solution, we use a Bayesian Monte-Carlo
methodology to estimate posterior probability density functions (PDFs)
of tectonic stresses, static fault friction, and pore fluid pressure.

@@

To quantify tectonic stresses on the Wenchuan earthquake faults, we
first calculate the topographic stress field in the upper crust
throughout eastern Tibet, then interpolate those stresses onto three
dimensional models of the faults taken from coseismic slip models.
Finally, we can calculate topographic shear and normal stresses on the
faults and compare those to the coseismic slip patterns.

@@

\subsection{Analytical description}\label{analytical-description}

We calculate the stress tensor field induced by topography throughout
eastern Tibet using methods developed by 
\citet{liuzoback1992}. They show that the topographic stress tensor
field can be determined by a convolution of a topographic loading
function with Green's functions describing the stresses in an elastic
halfspace due to a point load at the surface. In our notation, the
stress tensor field due to topography, $M(x,y,z)$, is given by

@@

where $G^B(x,y,z)$ are the Boussinesq solutions for stresses in a
halfspace due to a vertical point load on the surface (see Appendix A1)
and $F_v(x,y) = \rho g h(x,y)$. Note that $h(x,y)<0$ since $z<0$ is
depth. $M^C(x,y,z)$ is the component of the stress field due to the
mechanical coupling of the topography to the half-space, i.e.,
describing the lateral spreading forces in the rock above the halfspace,
and is given by

@@

where $G_i^C(x,y,z)$ are the Cerruti solutions for a horizontal point
source load in the $i$ direction on the halfspace surface (see Appendix
A1). The the horizontal loading functions are given by

@@

\begin{equation}
\begin{split}
F_{hor, \, yz}(x,y) = ( \rho g h(x,y) + \sigma_{yy}^B(x,y,0) + T_{yy} )\, \frac{\partial h}{ \partial y} \\
\quad + (M_{xy}^{B}(x,y,0) + T_{xy}) \frac{\partial h}{ \partial x}\; . 
\end{split}
\label{eqn:f_hor_yz}
\end{equation}

@@

Topography was taken from the CGIAR-CSI v.4 release
\citep{jarvis2008srtm} of the Shuttle Radar Topographic Mission
\citep{farr2007srtm} Digital Elevation Model (DEM) at 1 km nominal
resolution. The DEM was projected from native WGS84 geographic
coordinates to UTM zone 48N, decreasing the nominal horizontal
resolution to 851 m. We assume a Poisson halfspace, and Green's
functions for the Boussinesq and Cerruti point-source solutions were
calculated at regular points in large square, 2-D grid at each depth
considered, with the point-source centered in the grid (see Table
\ref{table:convo_params} for model parameters). A mask was applied to
each grid such that values outside the kernel radius (i.e.~the `corners'
of the array) were zero, so that the kernel is circular. The size of the
grid was chosen to be quite large to incorporate potential contributions
from the elevated topography throughout eastern Tibet. So that the
Green's functions and the topography were discretized on the same size
grid, we pad the Green's function array with zeros. Because of
singularities in the Green's functions at depth $z=0$, we use
$\sigma^B(x,y,z)$ with $z=851$ m, the shallowest level of our
calculations, in construction of the horizontal loading functions in
Equations \ref{eqn:f_hor_xz} and \ref{eqn:f_hor_yz}. Convolutions were
computed using a 2D fast Fourier transform. All calculations were
implemented in Python (v. 2.7.3) using IPython \citep{perez2007ipython},
NumPy (v. 1.7) \citep{oliphant2007numpy} and Pandas (v. 12)
\citep{mckinney2010}; additional statistical analysis was performed with
StatsModels \citep{seabold2010}. We created an open-source Python package
to calculate topographic stresses in a reasonably automated way, which
is available at \url{https://github.com/cossatot/halfspace}. The package
is being expanded to encompass a wide range of elastic stress and strain
solutions as time permits. All data and scripts for this particular
project are available at
\url{https://github.com/cossatot/wenchuan_topo_stress}.

@@

Topographic stresses on the Wenchuan faults are calculated on point sets
representing the faults taken from coseismic slip models. We use the
coseismic slip models of \citet{shen2009},
\citet{feng2010}, \citet{qi2011}, \citet{zhang2011},
and \citet{fielding2013}, and discard
points above 851 m below sea level, as this is above the depth at which
we compute $G^C(x,y,z)$. The six stress tensor components calculated at
the regular grid point are linearly interpolated to points describing
the faults. Because the fault points are completely surrounded by the
grid nodes at which topographic stresses were calculated and those nodes
are spaced \textless{}1 km apart, the fault points cannot be more than a
few hundred meters from the nearest grid node, so a higher order
interpolation is not necessary. We then project the topographic stress
tensor to fault normal stress, $\sigma_n^M$, down-dip shear stress,
$\tau_d^M$ and strike-slip shear stress, $\tau_s^M$ at each point in
describing the fault geometry.

@@

Topographic stresses on the Wenchuan faults are in the 1--10s of MPa
range (Figure~\ref{fig:lms_topo_stresses_rot}, ~\ref{fig:fault_stress_3d}). 
Stresses are highest in the southwest,
beneath the Pengguan massif (the highest topography of the Longmen Shan
front), and decrease to the northeast. $M_{zz}$ is typically larger,
though not substantially, than $M_{xx}$ or $M_{yy}$. Maximum horizontal
stress is not typically aligned with either cardinal horizontal
component, and is typically larger than $M_{zz}$ above 10 km. Maximum
$M_{zz}$ is near 80 MPa, on the southwestern Beichuan fault below the
high peaks of the Pengguan massif, except for in slip models containing
near-horizontal fault segments in the mid-crust, where $M_{zz}$ reaches
100 MPa. Vertical shear stresses ($M_{xz}$ and $M_{yz}$) are on the
order of 1 MPa, and horizontal shear stress ($M_{xy}$) is on the order
of 0.1 MPa.

\begin{figure}[t]
\centering
\includegraphics[width=20pc]{../figures/lms_topo_stresses_rot.pdf}
\caption{Horizontal topographic stresses in the Longmen Shan region at 5
km depth: black and red lines signify most and least compressive
horizontal stresses. Other symbols are the same as in Figure
1. Stresses shown are down-sampled from the
discretization used in the calculations by a factor of 6.}
\label{fig:lms_topo_stresses_rot}
\end{figure}

Because the compressive stresses are near equal, $M$ contains a large
isotropic component and a smaller shear component. Consequently, $M$
resolves on the Wenchuan faults with a large $\sigma^M_n$ (median of
about 40-60 MPa for each slip model) and much smaller $\tau^M_d$. The
median $\tau^M_d$ is about -3 to -6 MPa in each slip model, where values
less than zero indicate normal-sense shear, and $\tau^M_s$ median values
range from about -2 to 1 MPa, where values less than zero indicate
sinistral shear; fault models with positive median $\tau^M_s$ have broad
thrust flats at depth, where little slip occurred. On the steeper fault
segments (where most of the moment release occurred), topographic shear
stresses are typically normal-sinistral, as opposed to the dominant mode
of coseismic slip, which is reverse-dextral.

\begin{figure*}[htbp]
\centering
\includegraphics[width=40pc]{../figures/fault_stress_3d.pdf}
\caption{Southwest-looking views of topographic stresses and coseismic
slip on selected slip models. All models share same lateral extent, but
the perspectives for B and C are more inclined than A. (A) $\sigma_n^M$
(colors), slip magnitude (contours, 1 m interval) and hanging-wall
topography on the \citet{feng2010} model of the Beichuan
fault. Note the suppression of fault slip where normal stress is
highest, such as below the Pengguan massif (P). Fault and topography
share the same scale. No vertical exaggeration. (B) $\tau_d^M$ (colors)
and dip slip (contours, 1 m interval) on the \citet{qi2011}
`rough' slip model. (C) $\tau_s^M$ (colors) and strike slip (contours, 1
m interval) on the \citet{qi2011} `rough' slip model.}
\label{fig:fault_stress_3d}
\end{figure*}

In contrast to the steeper fault segments, much of the shallowly-dipping
fault segments (the Pengguan fault and flats at the base of the Beichuan
fault, where present) have $\tau^M$ in the direction of coseismic slip.
The full stress tensor $M$ is not significantly different in these
locations, but because of the low dip angle, $M_{zz}$ contributes more
significantly to $\sigma^M_n$ than to $\tau^M_d$, which is then
dominated by horizontal compression, leading to reverse-sense shear.
Additionally, the stresses caused by the Pengguan massif locally resolve
as right-lateral on these segments as well. Coseismic slip on these
fault patches is much lower than on the steeper Beichuan fault, where
the majority of slip occurred and which is topographically loaded in the
opposite shear sense.

Compellingly, similar patterns exist in the spatial distributions of
$\sigma^M_n$ and coseismic slip. Most obvious is the coincidence of
locally high $\sigma^M_n$ and locally low slip magnitude on the
southwestern Beichuan fault below the culmination of the Pengguan
massif, in an area of otherwise high slip (Figure
\ref{fig:fault_stress_3d}). These correlations exist for other fault
patches, but they are not as clear (Figure
\ref{fig:feng_slip_sig_n_scatter}). This raises the possibility that
$\sigma^M_n$ is capable of limiting slip once failure has occurred, and
may have implications for estimations of dynamic friction and the
completeness of stress drop during the earthquake. Preliminary analysis
of this is currently being performed, and will be described in a
forthcoming manuscript.

\begin{figure}%[htbp]
\centering
\includegraphics[width=20pc]{../figures/feng_slip_sig_n_scatter.pdf}
\caption{Coseismic slip magnitude and $\sigma^M_n$ on the four segments
of the \citet{feng2010} coseismic slip model. Trendlines are
L1 regressions and do not include points with no slip. NE B =
Northeastern Beichuan fault. C B = central Beichuan fault. SW B =
Southwestern Beichuan fault. P = Pengguan fault.}
\label{fig:feng_slip_sig_n_scatter}
\end{figure}

@@

We estimate the tectonic stress tensor field, $\mu$, and $\phi$
consistent with published coseismic slip models of the Wenchuan
earthquake using a Bayesian estimation. Bayesian estimation yields
posterior probability density functions of the model parameters \citep[e.g.,]
[]{mosegaard1995}. We first estimate posteriors of $T$ cosistent with
the coseismic slip models, and then estimate $\mu$ and $\phi$ consistent
with Mohr-Coulomb failure. The nature of Bayesian estimation allows us
to quantify both the relative likelihoods of model parameters and the
tradeoffs between them.

@@

where $M$ is described above, $T$ is the tectonic stress tensor, and $L$
is the lithostatic pressure tensor. $L$ is isotrapic, with diagonal
components equal to $\rho g z$. We assume that $T$ is latterally
homogeneous and only has horizontal stress components (i.e.,
$T_{xz} = T_{yz} = T_{zz} = 0$, with $T_{xx}$, $T_{yy}$ and $T_{xy}$
non-zero). We further assume that $T$ increases linearly with depth so
that the entire upper crust is near the critical failure envelope
\citep[e.g.,][]{townend2000}, are thus we parameterize the components of $T$
as scalars multiplied by lithostatic pressure.

@@

where $p(T)$ is the prior PDF (or \emph{priors}) of $T$, $p(D|T)$ is the
likelihood of observing the coseismic slip distribution $D$ given the
tectonic stresses $T$, and $p(T|D)$ is the posterior PDF of $T$ given
$D$, which is the solution to the inversion \citep[e.g.,][]{mosegaard1995}.
Due to the unknown proportionality in equation (\ref{eqn:bayes_rule}),
our posterior only gives likelihood of $T$ relative to the most likely
estiamte (MLE) \citep{tarantola2005}. We follow a Monte-Carlo strategy,
where samples of the prior PDF are retained as samples of the posterior
in proportion to $p(D|T)$ \citep[e.g.,][]{mosegaard1995}.

>>>>

@@

Stress is of fundamental importance to many processes in the earth. Both
the isotropic and deviatioric components of stress
exert control on the deformation state of the earth at any point in the
brittle regime. However, unlike other fundamental quantities such as
temperature, stress is typically difficult to measure \emph{in situ},
without drilling-based techniques. Therefore, stress is often treated in
a semi-quantitative manner, with an emphasis on directions and relative
magnitudes of the principal stresses, either locally or regionally
\citep[e.g.,][]{angelier1994}. These estimates of stress are commonly
derived from strain, for example from studies of earthquake focal
mechanisms \citep[e.g.,][]{michael1987} or of fault slip data
\citep[e.g.,][]{reches1987, medinaluna2013}.

@@

In this work, we investigate the effects of topographic stresses on the
faults that ruptured in the 2008 \emph{M}7.9 Wenchuan, China earthquake.
This earthquake is an ideal candidate for this study because it occurred at
the base of the Longmen Shan, one of the largest and steepest
escarpments on Earth (Figure~\ref{fig:lms_map}), and has a well-studied
coseismic slip distribution characterized by significant along-strike
variations in coseismic slip. Additionally, because the earthquake
occurred after \textasciitilde{}2000 years of seismic quiescence,
postseismic stresses in the lithosphere are likely to be negligible,
suggesting that the stress state on the faults at the time of rupture
can be approximated as the sum of topographic, lithostatic, and
accumulated tectonic stresses. We then use the Mohr-Coulomb failure
criterion to bracket the tectonic stress field, pore fluid pressure, and
static friction of the fault.

@@

Aspects of topographic stresses and their relevance to tectonics have
been studied for some time. \citet{jeffreys1924} noted that the
presence of high mountains is evidence that the Earth's crust can
support significant heterogeneous differential stresses over long time
periods. \citet{dalmayrac1981} and \citet{molnar1988} discussed how 
extensional deformation in
the high parts of orogens that is temporally coincident with contractional
deformation at the low-elevation margins of the orogen may be explained
by a spatially invariant, depth-integrated horizontal compressive stress
with spatially varying vertical stresses caused by topography.
\citet{richardson1994} exploited this
relationship in the central Andes to estimate horizontal tectonic
stresses through finite element modeling, and more recently 
\citet{copley2009} 
performed similar work in Albania. \citet{bollinger2004} showed how 
increased normal stress on the Main
Himalayan thrust due to loading of the high Himalayan massifs locally
suppressed microseismicity and increased fault locking. Meade and Conrad
\citet{meade2008} demonstrated that the increased weight of the
uplifting Andes influenced the Nazca-South America convergence rate.

Additionally, topographic stresses play a role in common models of
orogenic dynamics. For example, topographic loading is central to thin
viscous sheet models of lithospheric deformation \citep[e.g.,][]{birdpiper80,
flesch2010gpe}. Critical taper models for thrust or extensional
wedges \citep[e.g.,][]{dahlen1990, xiao1991} incorporate both variation in
vertical stress due to changing elevation as well as a shear stress
contributed by the slope of the wedge surface. Critical taper models
also incorporate the idea that in a growing wedge, progressively
increasing topographic stresses may eventually prevent continued slip on
a given fault plane, and strain will instead be transferred to the toe
of the thrust wedge where the stress state is more favorable. Of
particular relevance to our work are `fixed boundary' models of
gravitational collapse and spreading \citep{rey2001}, where an excess of
gravitational potential energy associated with the high topography and
thickened crust of eastern Tibet causes a transfer of rock to the
foreland through horizontal contraction at the rangefront \citep[e.g.,][]
{dewey1988, liuyang2003, copleymckenzie2007}. This process may be
aided by a weak (sub)horizontal structure at depth (such as a shear zone
or weak crustal channel) that is capable of transferring the vertical
and radial stresses from topography throughout an orogen to its margins,
leading to contraction there \citep[e.g.,][]{clark2005, burchfiel2008,
fleschbendick2012}.

The contributions of variable topography to the full stress field in the
elastic upper crust has been studied on smaller spatial scales. 
\citet{mctiguemei1981} and
\citet{savageswolfs1986} investigated the stress components from long,
symmetric ridges and showed how horizontal tension is induced under
ridge crests and horizontal compression is induced under valleys, mostly
due to shear stresses generated by slopes. Work in this vein was
continued by \citet{millerdunne1996} and \citet{martel2006}, who have 
focused on shallow rock fracturing
resulting from topographic stresses. \citet{liuzoback1992} investigated 
whether the topographic stresses
generated by the mountains around Cajon Pass (California, USA)
contributed to the observed left-lateral shear stress on a shallow
portion of the right-lateral San Andreas fault. In their study, they
developed methods for calculating the three dimensional elastic stress
tensor field arbitrary topography, whereas previous solutions were limited
to two dimensions and required topography to be mathematically-defined
(e.g., sinusoidal).

@@

Topographic stresses are only one component of the total stress field in
the crust \citep{molnar1988}. Coseismic slip in the Wenchaun earthquake
is due to the total accumulated stress on the faults, which also
includes components from lithostatic and tectonic stresses. (In the
present study, we do not consider stresses due to flexure \citep[e.g.,][]
{luttrell2007} or buoyancy \citep[e.g.,][]{luttrell2011}.) By quantifying
both the topographic and lithostatic stresses, we can use coseismic slip
models to solve for the tectonic stress assuming that (1) the total
shear stress on the fault is in the direction of coseismic slip \citep[e.g.,]
[]{angelier1994} and (2) the fault is at Mohr-Coulomb failure everywhere
that it slipped. If topographic stresses are significant and produce
shear in the direction of fault slip, then for given values of static
friction and pore fluid pressure, we can calculate the amount of
tectonic stress that can be added to the ambient stress field before the
faults should rupture; given limited acceptable ranges for friction and
fluid pressure, we are essentially able to place maximum constraints on
tectonic stress. Alternately, if topographic stresses work against
coseismic slip, for given friction and fluid pressures we can estimate
the minimum magnitudes of tectonic stresses necessary to overcome shear
and frictional resistance to slip. In a scenario with complex faulting
and topography, it may be possible to put tight bounds on minimum and
maximum magnitudes and directions of tectonic stresses. To account for
the non-uniqueness of the solution, we use a Bayesian Monte-Carlo
methodology to estimate posterior probability density functions (PDFs)
of tectonic stresses, static fault friction, and pore fluid pressure.

@@

To quantify tectonic stresses on the Wenchuan earthquake faults, we
first calculate the topographic stress field in the upper crust
throughout eastern Tibet, then interpolate those stresses onto three
dimensional models of the faults taken from coseismic slip models.
Finally, we calculate topographic shear and normal stresses on the
faults and compare those to the coseismic slip patterns.

@@

We calculate the stress tensor field induced by topography throughout
eastern Tibet using methods developed by 
\citet{liuzoback1992}. They show that the topographic stress tensor
field can be determined by a convolution of topographic loading
functions with Green's functions describing the stresses in an elastic
halfspace due to a point load at the surface. In our notation, the
stress tensor field due to topography is given by

@@

where $G^B(x,y,z)$ are the Boussinesq solutions for stresses in a
halfspace due to a vertical point load on the surface (see Appendix A1), 
$F_v(x,y) = \rho g h(x,y)$, and $h(x,y)$ is topography. 
Note that $h(x,y)<0$ since $z<0$ is
depth. $M^C(x,y,z)$ is the component of the stress field due to the
mechanical coupling of the topography to the half-space, i.e.,
describing the lateral spreading forces in the rock above the halfspace,
and is given by

@@

where $G_i^C(x,y,z)$ are the Cerruti solutions for a horizontal point
source load in the $i$ direction on the halfspace surface (see Appendix
A1). The horizontal loading functions are given by

@@

\begin{equation}
\begin{split}
F_{h, \, yz}(x,y) = ( \rho g h(x,y) + \sigma_{yy}^B(x,y,0) + T_{yy} )\, \frac{\partial h}{ \partial y} \\
\quad + (M_{xy}^{B}(x,y,0) + T_{xy}) \frac{\partial h}{ \partial x}\; . 
\end{split}
\label{eqn:f_hor_yz}
\end{equation}

@@

Topography was taken from the CGIAR-CSI v.4 release
\citep{jarvis2008srtm} of the Shuttle Radar Topographic Mission
\citep{farr2007srtm} Digital Elevation Model (DEM) at 1 km nominal
resolution. The DEM was projected from native WGS84 geographic
coordinates to UTM zone 48N, decreasing the nominal horizontal
resolution to 851 m. We assume a Poisson halfspace, and Green's
functions for the Boussinesq and Cerruti point-source solutions were
calculated at regular points in large 2-D grid at each depth
considered, with the point-source centered in the grid (see Table
\ref{table:convo_params} for model parameters). A mask was applied to
each of the discretized Green's functions such that values outside a 
radius (i.e. the `corners' of the array) were set to zero, yielding a
circular array. The size of the
grid was chosen to be quite large to incorporate potential contributions
from the elevated topography throughout eastern Tibet. So that the
Green's functions and the topography were discretized on the same size
grid, we pad the Green's function array with zeros. Because of
singularities in the Green's functions at $z=0$, we use
$\sigma^B(x,y,z)$ with $z=851$ m, the shallowest level of our
calculations, in construction of the horizontal loading functions in
Equations \ref{eqn:f_hor_xz} and \ref{eqn:f_hor_yz}. Convolutions were
computed using a 2D fast Fourier transform. All calculations were
implemented in Python (v. 2.7.3) using IPython \citep{perez2007ipython},
NumPy (v. 1.7) \citep{oliphant2007numpy} and Pandas (v. 12)
\citep{mckinney2010}; additional statistical analysis was performed with
StatsModels \citep{seabold2010}. We created an open-source Python package
to calculate topographic stresses in a reasonably automated way, which
is available at \url{https://github.com/cossatot/halfspace}. The package
is being expanded to encompass a wide range of elastic stress and strain
solutions as time permits. All data and scripts for this particular
project are available at
\url{https://github.com/cossatot/wenchuan_topo_stress}.

@@

Topographic stresses on the Wenchuan faults are calculated on point sets
representing the faults taken from coseismic slip models. We use the
coseismic slip models of \citet{shen2009},
\citet{feng2010}, \citet{qi2011}, \citet{zhang2011},
and \citet{fielding2013}, and discard
points above 851 m below sea level, as this is above the depth at which
we compute $G^C(x,y,z)$. The six stress tensor components calculated at
the regular grid points are linearly interpolated to points describing
the faults. Because the fault points are completely surrounded by the
grid nodes at which topographic stresses were calculated and those nodes
are spaced \textless{}1 km apart, the fault points cannot be more than a
few hundred meters from the nearest grid node, so a higher order
interpolation is not necessary. We then project the topographic stress
tensor to fault normal stress, $\sigma_n^M$, down-dip shear stress,
$\tau_d^M$ and strike-slip shear stress, $\tau_s^M$ at each point in
describing the fault geometry.

@@

Topographic stresses on the Wenchuan faults are on the order 1--10s MPa
(Figure~\ref{fig:lms_topo_stresses_rot}, ~\ref{fig:fault_stress_3d}). 
Stresses are highest in the southwest,
beneath the Pengguan massif (the highest topography of the Longmen Shan
front), and decrease to the northeast. $M_{zz}$ is typically larger,
though not substantially, than $M_{xx}$ or $M_{yy}$. Maximum horizontal
stress is not typically aligned with either cardinal horizontal
direction, and is typically larger than $M_{zz}$ above 10 km. Maximum
$M_{zz}$ is near 80 MPa, on the southwestern Beichuan fault below the
high peaks of the Pengguan massif, except for in slip models containing
near-horizontal fault segments in the mid-crust, where $M_{zz}$ reaches
100 MPa. Vertical shear stresses ($M_{xz}$ and $M_{yz}$) are on the
order of 1 MPa, and horizontal shear stress ($M_{xy}$) is on the order
of 0.1 MPa.

\begin{figure}[t]
\centering
\includegraphics[width=20pc]{../figures/lms_topo_stresses_rot.pdf}
\caption{Horizontal topographic stresses in the Longmen Shan region at 5
km depth: black and red lines signify most and least compressive
horizontal stresses, respectively. Other symbols are the same as in Figure
1. Stresses shown are down-sampled from the
discretization used in the calculations by a factor of six.}
\label{fig:lms_topo_stresses_rot}
\end{figure}

Because the compressive stresses are near equal, $M$ contains a large
isotropic component and a smaller deviatoric component. Consequently, $M$
resolves on the Wenchuan faults with a large $\sigma^M_n$ (median of
about 40-60 MPa for each slip model) and much smaller $\tau^M_d$. The
median $\tau^M_d$ is about -3 to -6 MPa in each slip model, where values
less than zero indicate normal-sense shear, and $\tau^M_s$ median values
range from about -2 to 1 MPa, where values less than zero indicate
sinistral shear; fault models with positive median $\tau^M_s$ have broad
thrust flats at depth, where little slip occurred. On the steeper fault
segments (where most of the moment release occurred), topographic shear
stresses are typically normal-sinistral, as opposed to the dominant mode
of coseismic slip, which is reverse-dextral.

\begin{figure*}[htbp]
\centering
\includegraphics[width=40pc]{../figures/fault_stress_3d.pdf}
\caption{Southwest-looking views of topographic stresses and coseismic
slip on selected slip models. All models share same lateral extent, but
the perspectives for B and C are more inclined than A. (A) $\sigma_n^M$
(colors), slip magnitude (contours, 1 m interval) and hanging-wall
topography on the \citet{feng2010} model of the Beichuan
fault. Note the suppression of fault slip where normal stress is
highest, such as below the Pengguan massif (P). Fault and topography
share the same scale, with no vertical exaggeration. (B) $\tau_d^M$ (colors)
and dip slip (contours, 1 m interval) on the \citet{qi2011}
`rough' slip model. (C) $\tau_s^M$ (colors) and strike slip (contours, 1
m interval) on the \citet{qi2011} slip model.}
\label{fig:fault_stress_3d}
\end{figure*}

In contrast to the steeper fault segments, much of the shallowly-dipping
fault segments (the Pengguan fault and flats at the base of the Beichuan
fault, where present) have $\tau^M$ in the direction of coseismic slip.
$M$ is not significantly different in these
locations, but because of the low dip angle, $M_{zz}$ contributes more
significantly to $\sigma^M_n$ than to $\tau^M_d$, which is then
dominated by horizontal compression, leading to reverse-sense shear.
The stresses caused by the Pengguan massif locally resolve
as right-lateral on these segments as well. Coseismic slip on these
fault patches is much lower than on the steeper Beichuan fault, where
the majority of slip occurred and which is topographically loaded in the
opposite shear sense.

Compellingly, similar patterns exist in the spatial distributions of
$\sigma^M_n$ and coseismic slip. Most obvious is the coincidence of
locally high $\sigma^M_n$ and locally low slip magnitude on the
southwestern Beichuan fault below the culmination of the Pengguan
massif, in an area of otherwise high slip (Figure
\ref{fig:fault_stress_3d}). These correlations exist for other fault
patches, but they are not as clear (Figure
\ref{fig:feng_slip_sig_n_scatter}). This raises the possibility that
topographic loading of these faults contributes to limiting coseismic
slip once failure has occurred, and
may have implications for estimations of dynamic friction and the
completeness of stress drop during the earthquake. Preliminary analysis
of this is currently being performed, and will be described in a
forthcoming manuscript.

\begin{figure}%[htbp]
\centering
\includegraphics[width=20pc]{../figures/feng_slip_sig_n_scatter.pdf}
\caption{Coseismic slip magnitude and $\sigma^M_n$ on the four segments
of the \citet{feng2010} coseismic slip model. Trendlines are
L1 regressions and do not include points with no slip. ``NE B'' =
Northeastern Beichuan fault. ``C B'' = central Beichuan fault. ``SW B'' =
Southwestern Beichuan fault. ``P'' = Pengguan fault.}
\label{fig:feng_slip_sig_n_scatter}
\end{figure}

@@

We estimate the tectonic stress tensor field, $\mu$, and $\phi$
consistent with published coseismic slip models of the Wenchuan
earthquake using a Bayesian estimation, resulting in posterior probability 
density functions of the model parameters. 
We first estimate posteriors of $T$ cosistent with
the coseismic slip models, and then estimate $\mu$ and $\phi$ consistent
with Mohr-Coulomb failure. The nature of Bayesian estimation allows us
to quantify both the relative likelihoods of model parameters and the
tradeoffs between them.

@@

where $M$ is described above, $T$ is the tectonic stress tensor, and $L$
is the lithostatic stress tensor. $L$ is isotropic, with diagonal
components equal to $\rho g z$. We assume that $T$ is latterally
homogeneous and only has horizontal stress components (i.e.,
$T_{xz} = T_{yz} = T_{zz} = 0$, with $T_{xx}$, $T_{yy}$ and $T_{xy}$
non-zero). We further assume that $T$ increases linearly with depth so
that the entire upper crust is near the critical failure envelope
\citep[e.g.,][]{townend2000}, are thus we parameterize the components of $T$
as scalars multiplied by lithostatic pressure, denoted as $T^\prime$

@@

where $p(T)$ is the prior PDF (or \emph{priors}) of $T$, $p(D|T)$ is the
likelihood of observing the coseismic slip distribution $D$ given the
tectonic stresses $T$, and $p(T|D)$ is the posterior PDF of $T$ given
$D$, which is the solution to the inversion \citep[e.g.,][]{mosegaard1995}.
Due to the unknown proportionality in equation (\ref{eqn:bayes_rule}),
our posterior only gives likelihood of $T$ relative to the most likely
estimate (MLE) \citep{tarantola2005}. We follow a Monte-Carlo strategy,
where samples of the prior PDF are retained as samples of the posterior
in proportion to $p(D|T)$ \citep[e.g.,][]{mosegaard1995}.
